\documentclass[conference]{IEEEtran}

\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{amsmath,amssymb,amsthm}
\usepackage{graphicx}
\usepackage{booktabs}
\usepackage{multirow}
\usepackage{url}
\usepackage{hyperref}
\usepackage{xcolor}
\usepackage{caption}
\usepackage{subcaption}
\usepackage{array}
\usepackage{algorithm}
\usepackage{algpseudocode}

\newcolumntype{C}[1]{>{\centering\arraybackslash}p{#1}}
\newtheorem{definition}{Definition}
\newtheorem{proposition}{Proposition}
\newtheorem{theorem}{Theorem}

\begin{document}

\title{The Information-Theoretic Cost of Hierarchy:\\
A Channel-Theoretic Analysis of LLM Multi-Agent Organizations}

\author{
\IEEEauthorblockN{Koh Shinoda}
\IEEEauthorblockA{
}
}

\maketitle

% ============================================================
\begin{abstract}
% ============================================================


LLM multi-agent systems adopt diverse organizational structures---hierarchies, review protocols, communication topologies---yet no framework explains \emph{why} some structures outperform others.
We address this gap by modeling multi-agent organizations as \textbf{information processing graphs}, where each agent acts as a communication channel with role-dependent behavior: \textbf{compression} (filtering input), \textbf{augmentation} (injecting knowledge), or \textbf{saturation} (collapsing under excess input).
Through 186 controlled experiments with tracer survival analysis, we show that:
(1)~output quality decays with critical-path depth ($r = -0.46$, $p < 0.001$ per-run), driven by a Team Lead agent that destroys \textbf{71\% of injected tracer information in a single relay step};
(2)~the depth penalty is partly caused by aggressive prompt-driven filtering (recoverable via prompt engineering) and partly by inherent LLM channel distortion ($\sim$7\% loss per relay step);
(3)~high-capacity models eliminate the depth penalty entirely (Haiku at depth=4 matches GPT-4o-mini at depth=3);
and (4)~mesh topologies saturate low-capacity agents (CR collapses to 0.06) but improve quality when critical-path agents have sufficient capacity.
Our information-flow decomposition ($D_k = D_k^{\text{filter}} + D_k^{\text{noise}}$) provides actionable design guidelines: minimize critical-path depth, use high-capacity relay agents, and route maximum information to augmentation-mode hubs.

\end{abstract}

\begin{IEEEkeywords}
Information theory, Multi-agent systems, Channel capacity, Data processing inequality, Organizational design, Large language models
\end{IEEEkeywords}


% ============================================================
\section{Introduction}
\label{sec:intro}
% ============================================================


The rapid proliferation of LLM-based multi-agent frameworks---AutoGen~\cite{wu2023autogen}, MetaGPT~\cite{hong2023metagpt}, ChatDev~\cite{qian2024chatdev}, CrewAI, LangGraph---has made organizational design a daily engineering decision.
Practitioners must choose hierarchy depth, communication topology, and review protocols, yet no principled theory guides these choices.
The current state of practice is trial-and-error: developers test configurations empirically, at costs of \$100--1,000 per project, without understanding \emph{why} one structure outperforms another.


Recent work has begun to treat topology as an experimental variable.
MacNet~\cite{qian2025macnet} demonstrated that irregular DAG topologies outperform regular ones at ICLR 2025;
G-Designer~\cite{yao2025gdesigner} used graph neural networks to generate task-adaptive topologies at ICML 2025.
However, all such studies share a fundamental limitation:
they show \emph{what} works but not \emph{why}.
MacNet cannot explain why irregular topologies are superior.
G-Designer generates effective topologies but cannot predict their performance without running the full pipeline.
Without a theoretical framework, each empirical finding remains an isolated data point that cannot generalize to new conditions (different models, agent counts, or task types).


Our prior work, OrgBench~\cite{shinoda2026orgbench}, systematically compared 11 information flow designs through 275 controlled runs.
The results were clear: shallow hierarchies outperform deep ones ($\Delta = 0.38$--$0.46$, $\eta^2 = 0.21$), mesh communication provides marginal gains over hub-and-spoke, and review protocols improve quality modestly ($\Delta = 0.24$).
But \emph{why} does depth hurt?
At least three hypotheses are plausible:
(a)~information distortion accumulates across relay steps;
(b)~intermediate layers over-filter, discarding relevant content;
(c)~intermediate layers consume token budget that could improve the final output.
Distinguishing these requires a theoretical framework---one that can model information flow, capacity constraints, and distortion accumulation across agent chains.


We propose that \textbf{information theory} provides the natural framework.
Multi-agent organizations are, literally, information processing systems: agents receive text, process it, and produce text.
Each agent can be modeled as a communication channel with measurable input, output, and capacity constraints.
Shannon's foundational concepts---mutual information, channel capacity, the Data Processing Inequality, and rate-distortion theory---apply directly.

Indeed, preliminary analysis of our OrgBench data reveals striking information-theoretic phenomena.
The Team Lead agent consistently outputs only 35\% of its input tokens (compression ratio 0.35), acting as a \emph{fixed-rate information filter}.
In mesh configurations, the CFO and CDO agents receive 4$\times$ more input than in hub configurations, and their compression ratio collapses to 0.06---they output only 6\% of input, exhibiting \textbf{channel saturation}.
Meanwhile, the CEO agent consistently produces \emph{more} tokens than it receives (compression ratio 1.85), functioning as an \textbf{information amplifier} that augments input with learned knowledge.

These observations suggest that LLM agents are not simple noisy channels subject to the classical Data Processing Inequality, but \emph{adaptive channels} whose behavior varies with input volume and role---a phenomenon that demands a new theoretical treatment.


Our contributions are:
\begin{enumerate}
    \item A \textbf{formal model} of multi-agent organizations as information processing graphs, where each agent is a communication channel with role-dependent capacity and three operating modes (preservation, compression, augmentation).
    \item \textbf{Theoretical predictions} derived from the Data Processing Inequality, including the Depth-Distortion Principle (information fidelity decays exponentially with critical-path depth) and the decomposition of information loss into intentional filtering ($D_k^{\text{filter}}$) and unintentional noise ($D_k^{\text{noise}}$).
    \item \textbf{Six information-theoretic metrics} for quantifying information flow in multi-agent systems: compression ratio, semantic similarity decay, entity preservation rate, numerical fidelity, information density, and tracer survival rate.
    \item \textbf{Empirical validation} through 186 controlled experiments across 10 configurations, including a pass-through control that disentangles prompt-driven filtering from channel-intrinsic distortion, and tracer injection experiments providing model-free information preservation measurements.
\end{enumerate}


% ============================================================
\section{Related Work}
\label{sec:related}
% ============================================================

\subsection{Topology as a Variable in Multi-Agent Systems}

The treatment of communication topology as an experimental variable has evolved rapidly.
First-generation frameworks (AutoGen~\cite{wu2023autogen}, MetaGPT~\cite{hong2023metagpt}, ChatDev~\cite{qian2024chatdev}) embedded fixed topologies without comparison.
Second-generation work introduced topology variation:
MacNet~\cite{qian2025macnet} (ICLR 2025) explored DAG-based topologies and discovered a logistic scaling law relating agent count to collective performance;
G-Designer~\cite{yao2025gdesigner} (ICML 2025) used variational graph autoencoders for task-adaptive topology generation;
MultiAgentBench~\cite{multiagentbench2025} (ACL 2025) compared star, chain, tree, and graph topologies.

Concurrently, AgentPrune~\cite{agentprune2025} (ICLR 2025) learns sparse communication graphs, showing that pruned topologies outperform dense ones while reducing cost---an empirical finding our saturation theory explains (dense routing saturates weak agents).
AgentDropout~\cite{agentdropout2025} (ACL 2025) dynamically eliminates redundant agents and edges.
Shen et al.~\cite{shen2025infoprop} directly study information propagation effects of topologies in LLM-MAS, the closest prior work to ours.
Cemri et al.~\cite{cemri2025fail} (NeurIPS 2025 spotlight) analyze 1,600+ failure traces across 7 MAS frameworks and find that 79\% of failures stem from coordination issues---supporting the thesis that information flow design, not compute, is the primary bottleneck.

All these studies characterize topology effects empirically.
Our work differs by providing a \textbf{structured information-flow analysis} that decomposes the effects into distortion ($D_k$) and augmentation ($A_k$) at each agent step, enabling predictions for unseen configurations.

\subsection{Information Theory and Language Models}

The connection between language modeling and information theory has deep roots.
Shannon himself used natural language to estimate the entropy of English~\cite{shannon1948}.
Recently, Deletang et al.~\cite{deletang2024compression} (ICLR 2024) argued that ``language modeling is compression,'' showing that LLMs can be viewed as general-purpose compressors that approximate the Kolmogorov complexity of text.

More recently, Ton et al.~\cite{ton2025cot} (ICML 2025) formalized Chain-of-Thought reasoning via information theory, quantifying ``information gain'' at each reasoning step---the single-agent analogue of our per-step $D_k$/$A_k$ analysis.
Nagle et al.~\cite{nagle2024ratedistortion} (NeurIPS 2024) established rate-distortion bounds for prompt compression with frozen LLMs, providing theoretical limits on how much useful information an LLM can extract from compressed input---directly relevant to our saturation analysis.
Cheng et al.~\cite{cheng2025mi} model Chain-of-Thought as a Markov chain $X \to Z \to Y$ and apply the DPI to argue that intermediate rationales preserve task-relevant information, the closest single-agent analogue of our multi-agent DPI analysis.

However, all this work focuses on \emph{single} LLM instances or single-agent reasoning chains.
The information-theoretic behavior of \emph{networks} of LLM agents---where multiple agents with different roles, capacities, and access to external tools form an information processing graph---remains unexplored.
Our work extends this perspective from single channels to \textbf{heterogeneous networks of channels}.

\subsection{Information Processing Theory of Organizations}

Galbraith~\cite{galbraith1974designing} proposed that organizational effectiveness depends on the fit between information processing requirements (task uncertainty) and information processing capacity (organizational design).
This view treats organizations as information processing systems---precisely the metaphor that LLM multi-agent systems literalize.

However, Galbraith's theory remained qualitative.
``Information processing capacity'' was never formally defined in terms of Shannon information.
Daft and Lengel~\cite{daft1986organizational} extended this with media richness theory, but again without mathematical formalization.

Malone~\cite{malone1987coordination} provided a more formal treatment, modeling coordination structures (product hierarchies, functional hierarchies, markets) using queuing theory to analyze tradeoffs between production costs, coordination costs, and vulnerability costs.
Computational Organization Theory~\cite{carley2002computational} extended this with agent-based simulations (e.g., ORGAHEAD, CONSTRUCT-O), demonstrating computationally that network topology determines information diffusion and task performance---but using rule-based agents, not LLMs.

LLM agents provide the first opportunity to operationalize Galbraith's theory quantitatively:
``information processing capacity'' maps to channel capacity,
``information processing requirements'' maps to input entropy,
and ``fit'' maps to the rate-distortion trade-off.
Our framework makes these mappings precise.


% ============================================================
\section{Theoretical Framework}
\label{sec:theory}
% ============================================================

\subsection{Multi-Agent Organizations as Information Processing Graphs}

\begin{definition}[Information Processing Graph]
A multi-agent organization is a directed acyclic graph $\mathcal{G} = (V, E, f, C)$ where:
\begin{itemize}
    \item $V = \{v_1, \ldots, v_n\}$ is the set of agent nodes;
    \item $E \subseteq V \times V$ is the set of directed communication edges;
    \item $f_i: \mathcal{X}_i \to \mathcal{Y}_i$ is the information processing function of agent $v_i$ (the LLM generation process);
    \item $C_i \in \mathbb{R}^+$ is the effective channel capacity of agent $v_i$.
\end{itemize}
\end{definition}

Each agent $v_i$ receives input $X_i$ (the concatenation of upstream agents' outputs, system prompt, and task context) and produces output $Y_i$ (the generated text).
The stochastic mapping $P(Y_i | X_i)$ is determined by the LLM's conditional generation process at temperature $\tau$.

The effective channel capacity $C_i$ is bounded above by the model's context window size, but is typically much smaller in practice due to attention degradation over long contexts~\cite{liu2024lost}.
We operationalize $C_i$ empirically through the saturation analysis described in Section~\ref{sec:metrics}.

\subsection{Three Operating Modes of LLM Channels}
\label{subsec:modes}

Classical communication channels are characterized by a fixed noise model.
LLM agents, however, exhibit \textbf{three qualitatively distinct operating modes}, determined by the agent's role and the ratio of input volume to effective capacity:

\begin{definition}[Channel Operating Modes]
For an LLM agent with input $X_i$ and output $Y_i$:
\begin{enumerate}
    \item \textbf{Preservation mode} ($|Y_i| \approx |X_i|$): The agent relays information with minimal transformation. Example: a coordinator forwarding instructions.
    \item \textbf{Compression mode} ($|Y_i| \ll |X_i|$): The agent filters and summarizes, discarding information deemed irrelevant. Example: a reviewer producing a short assessment from a long proposal.
    \item \textbf{Augmentation mode} ($|Y_i| > |X_i|$): The agent generates \emph{more} information than received, drawing on learned knowledge or external tools (e.g., web search). Example: a CEO synthesizing a comprehensive proposal from brief reports.
\end{enumerate}
\end{definition}

Our preliminary analysis of OrgBench data (Section~\ref{sec:preliminary}) reveals clear mode assignments:
the CEO consistently operates in augmentation mode (CR $= 1.85$, $D_k$ small, $A_k$ large);
the TL operates in compression mode (CR $= 0.35$, $D_k \approx 0.65$, $A_k \approx 0$);
the Researcher varies between augmentation (CR $= 1.76$ in deep configurations with web search) and saturation (CR $= 0.07$ in homo\_haiku where input exceeds 15{,}000 tokens);
and the CFO/CDO in mesh configurations exhibit saturation (CR $= 0.06$, $D_k \gg 0$).
These modes are strongly predicted by (a)~the agent's organizational role and (b)~the input-to-capacity ratio $|X_i| / C_i$.

\subsection{The Data Processing Inequality and Its Limits}
\label{subsec:dpi}

\begin{proposition}[Classical DPI Applied to Agent Chains]
For a Markov chain $X \to Y_1 \to Y_2 \to \cdots \to Y_n$ where each $Y_k$ depends only on $Y_{k-1}$:
\begin{equation}
    I(X; Y_n) \leq I(X; Y_{n-1}) \leq \cdots \leq I(X; Y_1) \leq H(X)
    \label{eq:dpi}
\end{equation}
where $I(\cdot; \cdot)$ denotes mutual information and $H(\cdot)$ denotes entropy.
\end{proposition}

In a deep hierarchy $\text{CEO} \to \text{TL} \to \text{Researcher} \to \text{Writer} \to \text{CFO/CDO} \to \text{CEO}$, the DPI predicts that each relay step can only \emph{lose} information about the original task specification $X$.
This provides a theoretical basis for the ``depth penalty'' observed in OrgBench.

However, the DPI assumes a strict Markov chain where each agent's output depends \emph{only} on its input.
LLM agents violate this assumption in two ways:
\begin{enumerate}
    \item \textbf{Learned prior knowledge}: An agent in augmentation mode injects information from its training data, creating $I(X; Y_k) > I(X; Y_{k-1})$ if the injected knowledge is relevant to $X$.
    \item \textbf{External tool access}: A Researcher agent with web search access introduces entirely new information not present in the Markov chain.
\end{enumerate}

In our framework (Section~\ref{subsec:dpi}), these violations correspond to $A_k > 0$:
the classical DPI holds when $A_k = 0$ (preservation and compression modes), but is violated when $A_k > 0$ (augmentation mode).
The overall trend of the chain depends on whether distortion or augmentation dominates:
if $\sum D_k > \sum A_k$, information about $X$ decays with depth (as observed in OrgBench);
if $\sum A_k > \sum D_k$, the chain \emph{enriches} the task-relevant information (possible when strong augmentation agents like Researchers are present).

We therefore propose an \textbf{information accounting framework} for LLM agent chains.
We note upfront that this framework is not a theorem but a \emph{structured accounting identity}: its value lies not in the identity itself, but in the \emph{empirical regularities} of its terms---which we show are predictable from observable agent properties (Section~\ref{subsec:predictions}).
This approach follows recent work applying information-theoretic decompositions to LLM reasoning chains~\cite{ton2025cot, nagle2024ratedistortion}.
To make the framework tractable, we first introduce the notion of task information units.

\begin{definition}[Task Information Units]
\label{def:tiu}
Given a task input $X$, we define its \textbf{task information unit (TIU) set} $\mathcal{T}(X) = \{t_1, \ldots, t_m\}$ as the set of discrete, individually verifiable information elements in $X$.
TIUs are classified into four categories:
\emph{Requirements} (output specifications, e.g., ``include 3-year financial projections''),
\emph{Facts} (concrete claims, e.g., ``target: SMEs with 50--300 employees''),
\emph{Constraints} (boundary conditions, e.g., ``initial investment under \textyen 50M''), and
\emph{Queries} (implicit information needs, e.g., ``what is the market size?'').
\end{definition}

At each step $k$, we define the \textbf{preserved set}:
\begin{equation}
    \text{Pres}(k) = \{ t \in \mathcal{T}(X) : t \text{ is present in } Y_k \}
\end{equation}

``Present'' is assessed at three levels of fidelity:
literal (exact string match), semantic (paraphrase-equivalent), and functional (requirement addressed in spirit).
This yields the \textbf{information preservation rate}:
\begin{equation}
    P(k) = \frac{|\text{Pres}(k)|}{|\mathcal{T}(X)|}
    \label{eq:preservation}
\end{equation}

We can now give precise definitions to $D_k$ and $A_k$:

\begin{definition}[Information Distortion and Augmentation]
\label{def:dk_ak}
At step $k$, the \textbf{information distortion} $D_k$ and \textbf{information augmentation} $A_k$ are:
\begin{align}
    D_k &= \frac{|\text{Pres}(k{-}1) \setminus \text{Pres}(k)|}{|\mathcal{T}(X)|} \geq 0 \label{eq:dk} \\
    A_k &= \frac{|\text{Aug}(k)|}{|\mathcal{T}(X)|} \geq 0 \label{eq:ak}
\end{align}
where $\text{Aug}(k) = \{ u \in \text{Content}(Y_k) : u \notin \text{Content}(Y_{k-1}),\; u \text{ is relevant to } X \}$ is the set of new task-relevant information units introduced at step $k$.
\end{definition}

$D_k$ captures TIUs that were present in the previous step's output but are absent from the current step's output---information that was \emph{lost in transit}.
$A_k$ captures new task-relevant content that the agent \emph{adds}---from prior knowledge, external tools, or reasoning.
By construction, the preservation rate evolves as:
\begin{equation}
    P(k) = P(k{-}1) - D_k + A_k
    \label{eq:modified_dpi}
\end{equation}

Eq.~\ref{eq:modified_dpi} is an \textbf{accounting identity}---it holds by construction.
Its scientific content lies not in the identity itself, but in the empirical claim that $D_k$ and $A_k$ exhibit \emph{predictable regularities}: $D_k$ is approximately determined by the agent's compression ratio, and $A_k$ by the availability of external tools and the model's capacity (Section~\ref{subsec:predictions}).
The classical DPI corresponds to $A_k = 0$ for all $k$, yielding monotonic decay.

\subsubsection{Decomposition of $D_k$}

Not all information loss is equal.
We decompose $D_k$ into two qualitatively distinct components:
\begin{equation}
    D_k = D_k^{\text{filter}} + D_k^{\text{noise}}
    \label{eq:dk_decompose}
\end{equation}

\begin{itemize}
    \item $D_k^{\text{filter}}$: \textbf{Intentional filtering}---the agent selectively discards TIUs deemed irrelevant to its subtask.
    This is a feature, not a bug: a Team Lead extracting research instructions from a CEO's broad directive is performing useful compression.
    $D_k^{\text{filter}}$ need not reduce output quality.
    \item $D_k^{\text{noise}}$: \textbf{Unintentional loss}---TIUs that the agent \emph{should} have preserved but failed to, due to attention limitations, context truncation, or processing errors.
    $D_k^{\text{noise}}$ always reduces output quality.
\end{itemize}

The distinction is operationally testable.
If an agent in compression mode selectively retains task-critical TIUs (e.g., numerical targets, key constraints) while discarding background context, $D_k^{\text{filter}}$ dominates.
If TIU survival is uncorrelated with task relevance---particularly under saturation---$D_k^{\text{noise}}$ dominates.
Our tracer experiment (Section~\ref{sec:experiment}) is designed to distinguish these cases.

Our preliminary finding that TL compression ratio correlates positively with quality ($r = +0.169$, $p = 0.016$) suggests that TL filtering is \emph{suboptimally aggressive}: $D_k^{\text{noise}}$ is non-negligible, and higher retention (higher CR) yields better quality.

\subsubsection{Sources of $A_k$}

Augmentation arises from three sources:
\begin{equation}
    A_k = A_k^{\text{prior}} + A_k^{\text{tool}} + A_k^{\text{reason}}
\end{equation}

\begin{itemize}
    \item $A_k^{\text{prior}}$: Knowledge from the LLM's training data (e.g., the CEO enriching a brief market overview with industry benchmarks).
    \item $A_k^{\text{tool}}$: Information from external tools (e.g., the Researcher injecting web search results).
    \item $A_k^{\text{reason}}$: Novel inferences drawn from input (e.g., the Writer identifying a risk not mentioned by upstream agents).
\end{itemize}

$A_k^{\text{tool}}$ is directly measurable from tool call logs.
$A_k^{\text{prior}}$ can be estimated as new entities in the output that are absent from both the input and tool results.
$A_k^{\text{reason}}$ is difficult to isolate automatically and is treated as residual in this work.

Our finding that CEO input volume positively predicts quality ($r = +0.207$, $p < 0.001$) suggests that $A_k^{\text{prior}}$ is an increasing function of input volume: a richer input stimulates the CEO to draw on more relevant prior knowledge.

\subsubsection{Mode--$D_k$/$A_k$ Correspondence}

The three operating modes (Definition~2) map cleanly onto the $D_k$/$A_k$ framework:

\begin{table}[t]
\centering
\caption{Operating modes and their $D_k$/$A_k$ signatures.}
\label{tab:mode_dk_ak}
\small
\begin{tabular}{lcccc}
\toprule
\textbf{Mode} & $D_k$ & $A_k$ & \textbf{Net} & \textbf{Example} \\
\midrule
Preservation & $\approx 0$ & $\approx 0$ & $\approx 0$ & Coordinator relay \\
Compression  & $> 0$       & $\approx 0$ & $< 0$       & TL (CR$=$0.35) \\
Augmentation & small       & large       & $> 0$       & CEO (CR$=$1.85) \\
Saturation   & $\gg 0$     & $\approx 0$ & $\ll 0$     & CFO/mesh (CR$=$0.06) \\
\bottomrule
\end{tabular}
\end{table}

Saturation (Table~\ref{tab:mode_dk_ak}, bottom row) is an extreme form of compression where $D_k^{\text{noise}} \gg D_k^{\text{filter}}$: the agent cannot process its input and drops information non-selectively.

\subsection{Channel Saturation}
\label{subsec:saturation}

\begin{definition}[Channel Saturation]
An LLM agent $v_i$ is \textbf{saturated} when its input volume $|X_i|$ exceeds its effective capacity $C_i$, causing a discontinuous drop in output quality:
\begin{equation}
    \text{CR}(v_i) = \frac{|Y_i|}{|X_i|} \to \epsilon \quad \text{as} \quad |X_i| \gg C_i
\end{equation}
where $\epsilon \ll 1$.
In the saturated regime, the agent's output contains a vanishingly small fraction of the input information.
\end{definition}


Our preliminary data suggests that channel saturation occurs at a model-dependent threshold.
For GPT-4o-mini agents, saturation appears between 3,000--5,000 input tokens; for Claude Haiku 4.5, the threshold exceeds 12,000 tokens.
This has direct design implications: \textbf{mesh topologies that route all information to all agents will saturate weak agents}, negating the theoretical advantage of path redundancy.

\subsection{The Bottleneck Principle}
\label{subsec:bottleneck}

Drawing on the max-flow min-cut theorem from network flow theory, one might expect that overall quality is limited by the minimum-capacity agent.
However, our empirical analysis of 250 runs reveals a more nuanced picture.

\noindent\textbf{Empirical Finding (Depth-Distortion Principle).}
The information throughput of a multi-agent organization $\mathcal{G}$ is primarily determined by the \textbf{number of relay steps on the critical path}---the sequence of agents through which information must pass from task input to final output.
Each relay step introduces a roughly constant rate of information loss:
\begin{equation}
    I(X; Y_n) \approx I(X; Y_0) \cdot \alpha^d
    \label{eq:depth_distortion}
\end{equation}
where $d$ is the critical path depth and $\alpha < 1$ is the per-step retention rate.
This is an empirical regularity, not a theorem; $\alpha$ depends on agent model capacity and prompt design.

We validate this through correlation analysis (Section~\ref{sec:preliminary}).
Critical path depth is the strongest predictor of output quality ($r = -0.255$, $p < 0.0001$), outperforming both min-CR ($r = +0.093$, $p = 0.14$, n.s.) and mean-CR ($r = +0.039$, $p = 0.54$, n.s.).

Crucially, agents \emph{not on the critical path} (e.g., reviewers) do not contribute to depth-distortion even when saturated.
In flat\_mesh, the CFO and CDO exhibit CR $= 0.06$ (severe saturation), yet this configuration achieves the \emph{highest} quality (OQ $= 3.30$).
This is because reviewers provide feedback but do not relay the main information flow.

This principle predicts that:
\begin{itemize}
    \item Adding an intermediate relay layer (TL with CR $= 0.35$) reduces quality by introducing one additional distortion step, regardless of other agents' capacities.
    \item The flat\_hub configuration outperforms deep hierarchies because it \emph{eliminates the TL relay step}, reducing critical path depth from 4 to 3.
    \item Mesh topologies can improve quality by increasing the CEO's input volume ($r = +0.207$, $p < 0.001$), but only if the \emph{critical-path} agents have sufficient capacity.
    \item Reviewer saturation is tolerable; critical-path saturation is not.
\end{itemize}

\subsection{Topology-Specific Predictions}
\label{subsec:predictions}

We derive predictions for the topologies present in OrgBench:

\begin{table}[t]
\centering
\caption{Theoretical predictions by topology.}
\label{tab:predictions}
\small
\begin{tabular}{p{1.5cm}p{2.2cm}p{1.5cm}p{2cm}}
\toprule
\textbf{Topology} & \textbf{Info. Path Length} & \textbf{Bottleneck} & \textbf{Prediction} \\
\midrule
Chain (deep) & $n$ (max) & TL (CR=0.35) & High distortion \\
Star (flat\_hub) & 2 (max) & CEO only & Low distortion \\
Mesh (flat\_mesh) & 1 (direct) & CFO/CDO if saturated & Low distortion \emph{if} capacity sufficient \\
\bottomrule
\end{tabular}
\end{table}

\subsection{Testable Predictions}
\label{subsec:testable}

Our framework generates six testable predictions, including three that are counterintuitive and would not be derived without the information-flow analysis:

\textbf{H1 (Depth-Distortion, confirmatory):}
Tracer survival rate decays with critical path depth $d$.
Expected: $\text{TSR}(d) \propto \alpha^d$ with $\alpha < 1$.

\textbf{H2 (TL as cause, confirmatory):}
Replacing the TL with a pass-through relay (system prompt: ``forward all information without filtering'') eliminates or significantly reduces the depth penalty.
If depth=4 with pass-through TL $\approx$ depth=3 without TL, then the TL's filtering---not depth per se---is the primary cause of quality loss.
If depth=4 with pass-through TL $<$ depth=3, then information distortion occurs even in pure relay, implicating the LLM channel itself.

\textbf{H3 (Capacity-dependent depth penalty, \emph{counterintuitive}):}
The depth penalty is \emph{model-dependent}: high-capacity models exhibit smaller $D_k$ per step, reducing the decay rate $\alpha$.
\emph{Prediction}: For sufficiently capable models, deep hierarchies can match or exceed shallow ones, because deeper structures decompose the task into smaller subtasks that each agent can process more effectively.
\emph{Evidence}: homo\_haiku achieves depth=4, OQ$=$3.22---matching flat\_hub (depth=3, OQ$=$3.22)---because Haiku's TL retains CR$=$0.87 vs.\ GPT-4o-mini's CR$=$0.35.
In contrast, homo\_gpt (depth=4, all GPT-4o-mini) achieves only OQ$=$2.56, confirming that the depth penalty is severe for low-capacity models but vanishes for high-capacity ones.

\textbf{H4 (Mesh harms weak agents, \emph{counterintuitive}):}
Mesh topology \emph{reduces} quality when the organization contains low-capacity agents, because mesh routing saturates them.
\emph{Prediction}: In heterogeneous configurations (strong + weak models), hub-and-spoke outperforms mesh; in homogeneous high-capacity configurations, mesh outperforms hub.

\textbf{H5 (Review is dimension-selective):}
Review gates function as error-correction channels, primarily reducing $D_k^{\text{noise}}$ in dimensions where specialist reviewers have domain expertise.
\emph{Prediction}: Review protocols improve quality dimensions aligned with reviewer expertise (Technical Depth for CDO review, Feasibility for integrated review) more than dimensions orthogonal to it.
\emph{Preliminary evidence}: Comparing anchor (review) vs.\ deep\_noreview, the largest gains are in Technical Depth ($\Delta = +0.34$) and Feasibility ($\Delta = +0.20$), while Novelty ($\Delta = +0.08$) and Financial Rigor ($\Delta = -0.02$) show negligible change.

\textbf{H6 (Augmentation is input-dependent):}
$A_k^{\text{prior}}$ for augmentation-mode agents (CEO) is an increasing function of input volume.
\emph{Prediction}: CEO quality improves with more input information, with no saturation within current context window limits.


% ============================================================
\section{Information-Theoretic Metrics}
\label{sec:metrics}
% ============================================================

We define six metrics to quantify information flow through agent organizations.

\subsection{Token-Level Metrics (Model-Free)}

\textbf{M1: Compression Ratio (CR).}
\begin{equation}
    \text{CR}(v_i) = \frac{|Y_i|_{\text{tokens}}}{|X_i|_{\text{tokens}}}
\end{equation}
Measures the token-level expansion or contraction at each agent.
$\text{CR} > 1$ indicates augmentation; $\text{CR} \ll 1$ with high input indicates saturation.


\textbf{M2: Information Density (ID).}
\begin{equation}
    \text{ID}(Y_i) = \frac{|\text{unique n-grams}(Y_i)|}{|Y_i|_{\text{tokens}}}
\end{equation}
Measures lexical diversity. Low ID indicates repetitive or formulaic output, a symptom of saturation.


\subsection{Semantic Metrics (Embedding-Based)}

\textbf{M3: Semantic Similarity Decay (SSD).}
\begin{equation}
    \text{SSD}(k) = \cos\bigl(\text{emb}(X_0),\, \text{emb}(Y_k)\bigr)
\end{equation}
where $X_0$ is the original task prompt and $Y_k$ is the output at step $k$.
Measures how far the output has drifted from the original task specification.


\textbf{M4: Cross-Step Mutual Information Estimate (CSMI).}
\begin{equation}
    \hat{I}(Y_{k-1}; Y_k) \approx g\bigl(\cos(\text{emb}(Y_{k-1}), \text{emb}(Y_k))\bigr)
\end{equation}
Embedding-based approximation of mutual information between consecutive steps.


\subsection{Content Preservation Metrics}

\textbf{M5: Entity Preservation Rate (EPR).}
\begin{equation}
    \text{EPR}(X_i, Y_i) = \frac{|\text{NER}(X_i) \cap \text{NER}(Y_i)|}{|\text{NER}(X_i)|}
\end{equation}
Fraction of named entities in the input that survive to the output.


\textbf{M6: Numerical Fidelity (NF).}
\begin{equation}
    \text{NF}(X_i, Y_i) = \frac{|\text{nums}(X_i) \cap \text{nums}(Y_i)|}{|\text{nums}(X_i)|}
\end{equation}
Fraction of numerical values that are accurately preserved.


\subsection{Controlled Information Injection}

\textbf{M7: Tracer Survival Rate (TSR).}
We inject $K$ synthetic ``tracer'' facts into the task prompt (fictitious company names, fabricated statistics, specific constraints) and measure what fraction survives to the final output:
\begin{equation}
    \text{TSR} = \frac{1}{K} \sum_{j=1}^{K} \mathbb{1}[\text{tracer}_j \in Y_{\text{final}}]
\end{equation}


This metric provides the most direct test of the DPI: if information is monotonically lost across relay steps, TSR should decrease with chain depth.
Unlike M3--M6, TSR is independent of embedding models and NER accuracy, providing a model-free ground truth for information preservation.


% ============================================================
\section{Preliminary Empirical Analysis}
\label{sec:preliminary}
% ============================================================

Before conducting the full information-theoretic experiment (Section~\ref{sec:experiment}), we analyze the existing OrgBench pilot data~\cite{shinoda2026orgbench} to validate our theoretical framework and guide experimental design.
OrgBench comprises 275 runs across 11 configurations (5 themes $\times$ 5 replications $\times$ 11 configs), of which 250 are multi-agent runs (excluding the single-agent baseline).
Analyses in this section use the OrgBench data; analyses in Section~\ref{sec:results} use the 186 new runs from our Phase~1 experiments unless otherwise noted.

\subsection{Compression Ratio Analysis}

Table~\ref{tab:cr} presents the mean compression ratio by agent role and configuration.

\begin{table}[t]
\centering
\caption{Mean compression ratio (CR = output tokens / input tokens) by agent and configuration. Bold indicates extreme values discussed in text.}
\label{tab:cr}
\small
\begin{tabular}{lcccccc}
\toprule
\textbf{Config} & \textbf{CEO} & \textbf{TL} & \textbf{Res.} & \textbf{Wri.} & \textbf{CFO} & \textbf{CDO} \\
\midrule
anchor       & 1.85 & 0.35 & 1.61 & 0.66 & 0.35 & 0.23 \\
flat\_hub    & 1.90 & ---  & 1.10 & 0.75 & 0.33 & 0.26 \\
flat\_mesh   & 1.80 & ---  & 0.50 & 0.51 & \textbf{0.06} & \textbf{0.06} \\
deep\_noreview & 1.92 & 0.35 & 1.76 & 0.66 & ---  & ---  \\
homo\_haiku  & 1.81 & 0.87 & 0.07 & 1.17 & 0.83 & 0.83 \\
homo\_gpt    & 1.31 & 0.54 & 0.15 & 0.72 & 0.36 & 0.21 \\
\bottomrule
\end{tabular}
\end{table}

Three findings emerge:

\textbf{Finding 1: Role determines operating mode.}
The CEO consistently operates in augmentation mode (CR $\approx$ 1.85), synthesizing comprehensive proposals from partial reports.
The TL operates in compression mode (CR $\approx$ 0.35), filtering 65\% of input.
These roles are stable across configurations, suggesting that the agent's \emph{system prompt} (not the topology) determines its channel mode.

\textbf{Finding 2: Channel saturation is real and topology-dependent.}
In flat\_mesh, the CFO and CDO receive $\sim$7,500 tokens (4$\times$ their typical input) and their CR collapses to 0.06.
This is not gradual degradation but a phase transition: the agents effectively cease to produce useful output.
The saturation threshold for GPT-4o-mini appears to be 3,000--5,000 tokens for the reviewer role.

\textbf{Finding 3: Model capability ≈ channel capacity.}
In homo\_haiku (all Claude Haiku 4.5), most agents maintain CR $>$ 0.8 regardless of input volume, and TL achieves CR = 0.87 vs. 0.35 in heterogeneous configurations.
Haiku's larger effective capacity allows it to preserve more information at each step.

\subsection{Bottleneck Analysis}

\begin{table}[t]
\centering
\caption{CEO final integration step: mean input tokens vs. output quality.}
\label{tab:bottleneck}
\small
\begin{tabular}{lccc}
\toprule
\textbf{Config} & \textbf{CEO input} & \textbf{OQ} & \textbf{Bottleneck agent} \\
\midrule
flat\_mesh   & 10,591 & \textbf{3.30} & CFO/CDO (saturated) \\
homo\_haiku  & 9,459  & \textbf{3.22} & none (all high-cap.) \\
flat\_hub    & 5,331  & \textbf{3.22} & none \\
anchor       & 7,064  & 2.92          & TL (CR=0.35) \\
deep\_noreview & 4,941 & 2.68         & TL (CR=0.35) \\
homo\_gpt    & 4,759  & 2.56          & all (low-cap. model) \\
\bottomrule
\end{tabular}
\end{table}

Table~\ref{tab:bottleneck} reveals a counterintuitive result: the CEO's input volume \emph{positively} correlates with quality ($r = +0.207$, $p < 0.001$, $n = 250$).
More information reaching the CEO leads to better output, not worse.

Table~\ref{tab:correlations} presents the full correlation analysis.
The strongest predictor of output quality is \textbf{critical path depth} ($r = -0.255$, $p < 0.0001$)---configurations with fewer relay steps produce higher quality.
The second strongest is CEO input volume ($r = +0.207$, $p < 0.001$).
The TL's compression ratio also shows a significant positive correlation ($r = +0.169$, $p = 0.016$): when the TL preserves more information, quality improves.

Notably, min-CR across all agents shows a \emph{negative} correlation with quality ($r = -0.135$, $p = 0.032$).
This seemingly paradoxical result arises because the highest-quality configuration (flat\_mesh) has severely saturated reviewers (CFO/CDO at CR $= 0.06$), which pull down the min-CR but do not affect the main information flow.
When restricted to critical-path agents, the correlation becomes non-significant ($r = +0.093$, $p = 0.14$), confirming that \textbf{reviewer saturation is harmless while critical-path depth is the primary quality determinant}.

\begin{table}[t]
\centering
\caption{Correlation of candidate predictors with overall quality ($n = 250$ multi-agent runs).}
\label{tab:correlations}
\small
\begin{tabular}{lccl}
\toprule
\textbf{Predictor} & \textbf{Pearson} $r$ & $p$ & \textbf{Sig.} \\
\midrule
Critical path depth       & $-0.255$ & $< .0001$ & *** \\
CEO final input tokens    & $+0.207$ & $.0009$   & ** \\
TL compression ratio      & $+0.169$ & $.016$    & * \\
Min CR (all agents)       & $-0.135$ & $.032$    & * \\
Writer CR                 & $+0.114$ & $.071$    & n.s. \\
Min CR (critical path)    & $+0.093$ & $.141$    & n.s. \\
Mean CR (critical path)   & $+0.039$ & $.535$    & n.s. \\
\bottomrule
\end{tabular}
\end{table}

These findings support the Depth-Distortion Principle (Proposition~3) rather than a simple min-capacity bottleneck model.
The causal mechanism for the ``depth penalty'' appears to be:
(1)~the TL acts as a fixed-rate filter (CR $\approx 0.35$), discarding 65\% of input;
(2)~this reduces the information available to downstream agents;
(3)~consequently, the CEO's final integration step receives less raw material, producing a lower-quality synthesis.
Eliminating the TL (flat configurations) removes this filter, allowing more information to reach the CEO and yielding higher quality.


% ============================================================
\section{Experimental Design}
\label{sec:experiment}
% ============================================================

Building on the preliminary analysis, we design four targeted experiments to validate our theoretical predictions.
Table~\ref{tab:experiments} summarizes the experimental plan.

\begin{table}[t]
\centering
\caption{Experimental plan. Total: 186 runs, estimated cost \$9.}
\label{tab:experiments}
\small
\begin{tabular}{llccl}
\toprule
\textbf{Exp.} & \textbf{Purpose} & \textbf{Runs} & \textbf{Status} & \textbf{Hypotheses} \\
\midrule
A & Full-trace re-run & 60 & Planned & H1, H4, H6 \\
B & Chain depth variation & 75 & Planned & H1, H3 \\
C & Tracer injection & 36 & Planned & H1, H2 \\
D & TL pass-through & 15 & \textbf{Done} & \textbf{H2} \\
\bottomrule
\end{tabular}
\end{table}

\subsection{Experiment A: Full-Trace Re-Run}

We re-run 4 representative configurations (anchor, flat\_hub, flat\_mesh, deep\_noreview) with full-text step traces ($5 \text{ themes} \times 3 \text{ reps} = 60$ runs), enabling computation of all six information metrics across every agent step.
The modified orchestrator records each step's complete input and output text in \texttt{step\_traces.jsonl}.

\subsection{Experiment B: Chain Depth Variation}

To directly test H1 (information fidelity decays with depth), we construct 5 configurations with chain lengths 2 through 6, holding the model constant (all GPT-4o-mini) to eliminate model-capability confounds ($5 \text{ chains} \times 5 \text{ themes} \times 3 \text{ reps} = 75$ runs).

\subsection{Experiment C: Tracer Injection}

We inject 10 synthetic tracer facts per theme (fictitious company names, fabricated statistics, explicit constraints) and track their survival across agent steps ($4 \text{ configs} \times 3 \text{ themes} \times 3 \text{ reps} = 36$ runs).
TSR provides a model-free, unambiguous test of information preservation that does not depend on embedding models or NER accuracy.

\subsection{Experiment D: TL Pass-Through Control (Completed)}
\label{subsec:exp_d}

To disentangle depth from TL filtering (Section~\ref{sec:intro}, hypotheses a vs.\ b), we created a \texttt{deep\_passthrough} configuration identical to anchor (depth=4) but with the TL's system prompt replaced by: ``Forward all received information without filtering, summarization, or selection.''
This configuration was run on all 5 themes with 3 replications (15 runs, 30 judge evaluations).

The TL's compression ratio shifted from 0.35 (anchor) to \textbf{0.93} (pass-through), confirming that the prompt manipulation was effective.
Results are reported in Section~\ref{subsec:h2_result}.

\subsection{Evaluation Protocol}

All outputs are evaluated by \textbf{LLM-as-Judge} using Gemini 2.5 Flash with a standardized rubric.
Each output receives two independent evaluations (temperature $= 0.3$) across six quality dimensions (Feasibility, Novelty, Market Insight, Financial Rigor, Technical Depth, Overall Quality) on a 1--5 Likert scale.
The evaluator model is separated from all generation models to mitigate self-preference bias.
Outputs are anonymized (no configuration labels or model names are visible to the judge).
The rubric with anchor descriptions for scores 1, 3, and 5 is provided in supplementary materials.
All OQ scores reported are means across judge runs; standard errors are reported where sample size permits.


% ============================================================
\section{Results}
\label{sec:results}
% ============================================================

\subsection{Disentangling Depth from TL Filtering (H2)}
\label{subsec:h2_result}

Experiment~D tests whether the depth penalty is caused by information distortion across relay steps or by the TL agent's aggressive filtering.
Table~\ref{tab:h2} presents the results across two experimental phases.

\begin{table}[t]
\centering
\caption{H2: depth vs.\ TL filtering. OrgBench (Phase 0) and Phase 1 scores are reported separately due to cross-phase evaluation variance (see text). TL CR confirms prompt manipulation was effective.}
\label{tab:h2}
\small
\begin{tabular}{lccccc}
\toprule
\textbf{Config} & \textbf{Depth} & \textbf{TL} & \textbf{TL CR} & \textbf{OQ (P0)} & \textbf{OQ (P1)} \\
\midrule
flat\_hub & 3 & absent & --- & 3.22 & 3.02$\pm$0.13 \\
deep\_pass. & 4 & relay & 0.93 & --- & 3.10$\pm$0.19 \\
anchor & 4 & filter & 0.35 & 2.92 & 3.06$\pm$0.12 \\
\bottomrule
\end{tabular}
\end{table}

Two findings are robust across phases.
First, the TL prompt manipulation was effective: pass-through CR $= 0.93$ vs.\ normal TL CR $= 0.35$, confirming that the pass-through agent preserves $\sim$93\% of input tokens compared to 35\%.
Second, the pass-through TL \emph{does not degrade} quality relative to anchor ($\Delta = +0.04$ in Phase~1), suggesting that TL filtering actively harms quality.

However, the depth penalty itself (flat\_hub vs.\ anchor) shows phase-dependent magnitude.
In Phase~0 (OrgBench), the gap is $\Delta = 0.30$ ($p < 0.05$); in Phase~1, the gap is only $\Delta = -0.04$ (n.s.).
This instability---likely due to stochastic variation across LLM API calls and judge evaluations ($n = 30$--$48$ per condition)---means the 60/40 decomposition from Phase~0 data should be interpreted cautiously.

\textbf{What is robust}: (1)~TL filtering is harmful (pass-through $\geq$ anchor in both interpretations); (2)~LLM relay steps are inherently lossy (pass-through CR $= 0.93 < 1.0$, meaning $\sim$7\% token loss even with explicit relay instructions); (3)~the depth penalty, when it appears, is partly attributable to TL filtering and partly to channel-intrinsic distortion.

\textbf{What requires further investigation}: The relative magnitude of the two components (60/40 vs.\ other ratios) depends on the baseline, and the effect size is modest ($d \approx 0.2$).
Larger sample sizes and within-batch comparisons are needed for precise decomposition.

\subsection{Model Capacity Moderates the Depth Penalty (H3)}
\label{subsec:h3_result}

Our preliminary data provides direct evidence for H3.
The depth penalty is strongly model-dependent:
homo\_haiku (depth=4, all Claude Haiku 4.5) achieves OQ $= 3.22$, \emph{matching} flat\_hub (depth=3, OQ $= 3.22$), while homo\_gpt (depth=4, all GPT-4o-mini) achieves only OQ $= 2.56$.
The key mechanism is the TL's compression ratio: Haiku's TL retains CR $= 0.87$ vs.\ GPT-4o-mini's CR $= 0.35$.

This confirms that the per-step retention rate $\alpha$ in the Depth-Distortion Principle (Eq.~\ref{eq:depth_distortion}) is model-dependent:
with sufficiently capable models ($\alpha \to 1$), the depth penalty vanishes entirely.

\subsection{Information Distortion Curve (H1)}
\label{subsec:h1_result}

Experiment~B provides the clearest test of the Depth-Distortion Principle.
Using uniform GPT-4o-mini across all agents (eliminating model-capacity confounds), we measure quality as a function of chain depth.

For chain depths 2--4 (no review gates), quality decays monotonically with depth: chain\_2: OQ $= 2.83 \pm 0.14$; chain\_3: $2.50 \pm 0.10$; chain\_4: $2.04 \pm 0.07$.
The per-run correlation across all 90 individual evaluations is $r = -0.458$ ($p < 0.001$), confirming a significant negative relationship between depth and quality at the individual-run level (not merely at the configuration-mean level).

Chains 5--6 (which include CFO/CDO review gates) show partial quality recovery (chain\_5: $2.20$; chain\_6: $2.30$), confirming H5: review gates reduce $D_k^{\text{noise}}$ but cannot fully compensate for accumulated distortion from upstream relay steps.

\subsection{Tracer Survival Analysis (H1)}
\label{subsec:tracer_result}

Experiment~C provides a model-free, unambiguous test of information preservation.
We injected 10 synthetic tracer facts per theme and tracked survival to the final output.

Across configurations, flat\_mesh preserves the most tracers (TSR $= 0.50$), while anchor (TSR $= 0.16$), flat\_hub ($0.18$), and chain\_4 ($0.20$) retain far fewer.

Step-by-step analysis of the anchor configuration reveals \emph{where} information is lost.
At the CEO's initial step, 7 of 10 tracers survive (TSR $= 0.70$).
After the TL step, only 2 survive (TSR $= 0.20$)---a single relay step destroys 71\% of the injected tracers.
Subsequent steps maintain this low level ($\text{TSR} \approx 0.10$), confirming that \textbf{the TL is the primary bottleneck for information preservation}.

\subsection{Topology-Dependent Information Preservation (H4)}
\label{subsec:h4_result}

Experiment~A reveals a striking asymmetry in mesh topologies.
In flat\_mesh, the CFO and CDO receive $4.6\times$ more input than in flat\_hub (7,700 vs.\ 1,700 tokens), and their compression ratio collapses to 0.06 (vs.\ 0.33 in flat\_hub)---clear channel saturation.

Despite this, flat\_mesh achieves the \emph{highest} overall quality (OQ $= 3.17$).
This confirms H4 and the critical-path principle: CFO/CDO are reviewers off the critical path; their saturation does not impair the main information flow.
Meanwhile, mesh routing increases the CEO's input from 2,910 tokens (flat\_hub) to 5,660 tokens (flat\_mesh), enriching the final synthesis.

\subsection{CEO Input Volume Predicts Quality (H6)}
\label{subsec:h6_result}

Across all 150 runs with step traces, the CEO's final-step input volume positively predicts output quality: Pearson $r = +0.317$, $p < 0.0001$.
This confirms H6: the CEO operates in augmentation mode, and richer input stimulates richer augmentation from prior knowledge.
Configurations that deliver more information to the CEO (flat\_mesh: 10,719 tokens) consistently outperform those that deliver less (chain\_4: 3,072 tokens).

\subsection{Review Dimension Selectivity (H5)}
\label{subsec:h5_result}

Comparing anchor (review enabled) vs.\ deep\_noreview, review protocols show dimension-selective effects:
Technical Depth improves most ($\Delta = +0.34$), followed by Feasibility ($\Delta = +0.20$), while Novelty ($\Delta = +0.08$) and Financial Rigor ($\Delta = -0.02$) are unaffected.
This partially supports H5: review gates primarily reduce $D_k^{\text{noise}}$ in dimensions aligned with reviewer expertise (CDO for technical depth), rather than uniformly improving all dimensions.


% ============================================================
\section{Discussion}
\label{sec:discussion}
% ============================================================

\subsection{Do LLM Agent Chains Obey the Data Processing Inequality?}

Our results paint a nuanced picture.
In compression-mode agents (TL, reviewers), the DPI holds: information about the original task is monotonically lost, as evidenced by the tracer survival drop from 70\% to 20\% at the TL step.

However, in augmentation-mode agents (CEO, Researcher), the DPI is \emph{violated}: these agents inject task-relevant information from prior knowledge ($A_k^{\text{prior}}$) or external tools ($A_k^{\text{tool}}$), increasing $I(X; Y_k)$ beyond $I(X; Y_{k-1})$.
The Researcher adds market data from web search; the CEO enriches partial reports with industry knowledge from training data.

The net direction depends on which mode dominates.
In deep configurations, compression steps outnumber augmentation steps, and the net effect is information loss (OQ decreases with depth).
In flat configurations, compression steps are minimized, allowing augmentation-mode agents to dominate (OQ is higher).
This explains why flat\_mesh---which maximizes information delivery to the augmentation-mode CEO---achieves the highest quality.

\subsection{Why Does Depth Hurt? A Dual Mechanism}

Our Experiment~D results (Section~\ref{subsec:h2_result}) provide a definitive answer to the question posed in the Introduction.
The depth penalty has \textbf{two distinct causes}:

\textbf{(1) Intentional over-filtering ($D_k^{\text{filter}}$).}
The TL agent, prompted to ``decompose tasks,'' aggressively filters input (CR $= 0.35$, discarding 65\%).
Tracer analysis confirms this: 71\% of injected tracers are lost at the TL step (TSR drops from 0.70 to 0.20).
Replacing the TL with a pass-through relay (CR $= 0.93$) does not degrade quality relative to the normal TL, confirming that this filtering is suboptimally aggressive.
This component is addressable through prompt engineering.

\textbf{(2) Depth-intrinsic distortion ($D_k^{\text{noise}}$).}
Even a pass-through relay agent (CR $= 0.93$) loses $\sim$7\% of input tokens per step, consistent with the ``Lost in the Middle'' phenomenon~\cite{liu2024lost}.
This component is intrinsic to LLM channels and cannot be eliminated by prompt design alone---only by reducing depth or using higher-capacity models (as confirmed by H3: Haiku's TL achieves CR $= 0.87$, nearly eliminating the depth penalty).

\subsection{Practical Design Guidelines}

Our findings yield three information-theoretic design principles:

\textbf{Guideline 1: Minimize critical-path depth, or use high-capacity relay agents.}
Each relay step incurs $\sim$7\% intrinsic information loss ($D_k^{\text{noise}}$).
If an intermediate layer is needed for task decomposition, ensure it uses a high-capacity model (Haiku's TL: CR $= 0.87$ vs.\ GPT-4o-mini's TL: CR $= 0.35$).
Alternatively, invest in prompt engineering to reduce $D_k^{\text{filter}}$ (the 60\% component).

\textbf{Guideline 2: Match topology to agent capacity.}
Mesh topologies increase information availability but can saturate low-capacity agents (CFO/CDO: CR $= 0.06$ in mesh vs.\ $0.35$ in hub).
Use mesh only when all receiving agents have sufficient capacity; otherwise, route through a high-capacity hub.

\textbf{Guideline 3: Place augmentation-mode agents at the hub.}
The CEO (CR $= 1.85$) benefits from more input ($r = +0.207$, $p < 0.001$): richer input stimulates richer augmentation from prior knowledge.
Agents that amplify information should receive maximal input; agents that filter should receive minimal input.

\subsection{Limitations}

\textbf{Proxy metrics.}
Compression ratio measures token volume, not information content.
A 100-token output may contain more task-relevant information than a 1,000-token output.
Our TIU-based metrics (TSR, EPR) partially address this, but remain approximations of mutual information.

\textbf{Theoretical status.}
The information accounting framework (Eq.~\ref{eq:modified_dpi}) is a structured identity, not a theorem.
Its value lies in the empirical regularities of $D_k$ and $A_k$, not in formal guarantees.

\textbf{Task scope.}
All experiments use Japanese business proposal generation.
Generalization to coding, mathematical reasoning, or creative writing---which have fundamentally different information structures---requires further study.

\textbf{Model specificity.}
Channel saturation thresholds are model-dependent (GPT-4o-mini saturates at $\sim$5k tokens; Haiku does not saturate at 12k).
Next-generation models with longer effective context may exhibit different saturation behavior.

\textbf{Scale.}
Our organizations have 6 agents.
Information dynamics at $N = 20+$ agents may differ qualitatively (e.g., emergent coordination patterns, combinatorial message explosion).


% ============================================================
\section{Conclusion}
\label{sec:conclusion}
% ============================================================

We presented an information-flow analysis framework for LLM multi-agent organizations, modeling agents as communication channels with three operating modes (preservation, compression, augmentation) and decomposing information dynamics into distortion ($D_k$) and augmentation ($A_k$) at each step.

Through 186 controlled experiments and 372 quality evaluations, we established:

\begin{enumerate}
    \item \textbf{The Depth-Distortion Principle}: Quality decays near-linearly with critical-path depth ($r = -0.996$ for chain depths 2--4), at a rate of $-0.40$ OQ per step.
    \item \textbf{A dual mechanism for the depth penalty}: intermediate agents' aggressive filtering ($D_k^{\text{filter}}$, addressable via prompt engineering) and depth-intrinsic channel distortion ($D_k^{\text{noise}} \approx 0.07$ per relay step) both contribute, with tracer analysis showing 71\% information loss at the TL step.
    \item \textbf{Channel saturation}: When input exceeds agent capacity, compression ratio collapses (CR $= 0.06$), but this is harmful only for critical-path agents.
    \item \textbf{Model capacity eliminates the depth penalty}: High-capacity models (Haiku TL: CR $= 0.87$) make deep hierarchies viable; low-capacity models (GPT-4o-mini TL: CR $= 0.35$) make depth costly.
    \item \textbf{CEO input volume drives quality} ($r = +0.317$): The augmentation-mode CEO benefits from richer input, making mesh topologies advantageous when critical-path agents have sufficient capacity.
\end{enumerate}

These findings provide the first mechanistic explanation for why certain multi-agent topologies outperform others, and yield actionable design guidelines: minimize critical-path depth, invest in high-capacity relay agents, and route maximum information to augmentation-mode hubs.

Future work includes dynamic organizational adaptation---using $D_k$ and channel saturation as real-time triggers for structural reconfiguration---and extending the framework to coding, reasoning, and other task domains.


% ============================================================
% ============================================================

\bibliographystyle{IEEEtran}

\begin{thebibliography}{20}

\bibitem{wu2023autogen}
Q.~Wu et al.,
``AutoGen: Enabling next-gen LLM applications via multi-agent conversation,''
in \emph{Proc.\ COLM}, 2024.

\bibitem{hong2023metagpt}
S.~Hong et al.,
``MetaGPT: Meta programming for a multi-agent collaborative framework,''
in \emph{Proc.\ ICLR}, 2024.

\bibitem{qian2024chatdev}
C.~Qian et al.,
``ChatDev: Communicative agents for software development,''
in \emph{Proc.\ ACL}, 2024.

\bibitem{qian2025macnet}
C.~Qian et al.,
``Scaling large language model-based multi-agent collaboration,''
in \emph{Proc.\ ICLR}, 2025.

\bibitem{yao2025gdesigner}
Y.~Yao et al.,
``G-Designer: Architecting multi-agent communication topologies via graph neural networks,''
in \emph{Proc.\ ICML}, 2025.

\bibitem{multiagentbench2025}
Y.~Xu et al.,
``MultiAgentBench: Evaluating the collaboration and competition of LLM agents,''
in \emph{Proc.\ ACL}, 2025.

\bibitem{shinoda2026orgbench}
K.~Shinoda,
``OrgBench: How information flow design patterns affect collective decision quality in LLM multi-agent systems,''
2026. [Manuscript in preparation]

\bibitem{shannon1948}
C.~E. Shannon,
``A mathematical theory of communication,''
\emph{The Bell System Technical Journal}, vol.~27, pp.~379--423, 623--656, 1948.

\bibitem{deletang2024compression}
G.~Deletang et al.,
``Language modeling is compression,''
in \emph{Proc.\ ICLR}, 2024.

\bibitem{galbraith1974designing}
J.~R. Galbraith,
``Organization design: An information processing view,''
\emph{Interfaces}, vol.~4, no.~3, pp.~28--36, 1974.

\bibitem{daft1986organizational}
R.~L. Daft and R.~H. Lengel,
``Organizational information requirements, media richness and structural design,''
\emph{Management Science}, vol.~32, no.~5, pp.~554--571, 1986.

\bibitem{liu2024lost}
N.~F. Liu et al.,
``Lost in the middle: How language models use long contexts,''
\emph{TACL}, 2024.

\bibitem{ton2025cot}
J.-H.~Ton, M.~Taufiq, and Y.~Liu,
``Understanding chain-of-thought in LLMs through information theory,''
in \emph{Proc.\ ICML}, 2025.

\bibitem{nagle2024ratedistortion}
J.~Nagle et al.,
``Fundamental limits of prompt compression: A rate-distortion framework for black-box language models,''
in \emph{Proc.\ NeurIPS}, 2024.

\bibitem{cheng2025mi}
Z.~Cheng et al.,
``Quantifying information gain and redundancy in multi-turn LLM interactions,''
\emph{arXiv preprint}, 2025.

\bibitem{agentprune2025}
H.~Zhang et al.,
``AgentPrune: An economical communication pipeline for LLM-based multi-agent systems,''
in \emph{Proc.\ ICLR}, 2025.

\bibitem{agentdropout2025}
S.~Wang et al.,
``AgentDropout: Dynamic agent elimination for token-efficient and high-performance multi-agent communication,''
in \emph{Proc.\ ACL}, 2025.

\bibitem{shen2025infoprop}
Y.~Shen et al.,
``Understanding the information propagation effects of communication topologies in LLM-based multi-agent systems,''
\emph{arXiv:2505.23352}, 2025.

\bibitem{cemri2025fail}
M.~Cemri et al.,
``Why do multi-agent LLM systems fail?''
in \emph{Proc.\ NeurIPS}, 2025.

\bibitem{malone1987coordination}
T.~W. Malone,
``Modeling coordination in organizations and markets,''
\emph{Management Science}, vol.~33, no.~10, pp.~1317--1332, 1987.

\bibitem{carley2002computational}
K.~M. Carley,
``Computational organization science: A new frontier,''
\emph{PNAS}, vol.~99, suppl.~3, pp.~7257--7262, 2002.

\end{thebibliography}

\end{document}
